---
title: "Bald eagle in British Columbia - Spatial Exploration"
author: "Shayla Tran, Matthew Angoh & Kyle Deng"
date: "2024-04-17"
output: html_document
---

There is a total of 104,941 occurrences recorded of Bald eagle or *Haliaeetus leucocephalus (Linnaeus, 1766)* within Canada between 1848 and 2024. This data is directly downloaded from the source GBIF.

## 1) Initial Data Inspection and Data Exploration

Here, we load in the `BC_Covariates.Rda` data. We will also read in the `eagles_bc.csv` dataset which contains 5000 sample occurrences of Bald eagle.

```{r, include=FALSE, message=FALSE}
# Load necessary libraries
library(tidyverse)
library(spatstat)
library(sf)
library(polyCub)
```

```{r, warning=FALSE, message=FALSE}
load("data/BC_Covariates.Rda") # BC covariate data with window
occurrence_eagles <- read_csv("data/eagles_bc.csv") # Reading in the occurrences data
```

Creating the `ppp` object and plotting the Bald eagle occurrences into the BC window.

```{r}
window <- as.owin(DATA$Window) # Create an owin object with the window in the covariates dataset so it can be used to create the ppp object later

# Create the ppp object
eagles_ppp <- ppp(x = occurrence_eagles$X, y = occurrence_eagles$Y, window = window) # Obtain a ppp object with the coordinate and window data 

# Plot the ppp object
plot(eagles_ppp,
     cols = 4,
     pch = 16,
     main = "Eagles occurrence data in BC",
     cex = 0.5)
```

From the initial plot, we can see that the eagles tend to be near the southern regions and there are some densities along the coast and within the interior region.

The BC covariates provided are:

* HFI (Human Foot Index): influence of humans on the land
* Elevation: land elevation
* Forest: forested areas
* Dist_water: distance to water sources or bodies of water

To explore these covariates, we will create plots that show the eagle occurrences along with the covariate data within the window. Each covariate image will be split into 3 elevation classes (for now).

```{r, hide=TRUE}
# Plotting the covariates (for comparisons later)
plot(DATA$Elevation, main = "Elevation")
plot(DATA$Forest, main = "Forest Coverage")
plot(DATA$HFI, main = "Human Foot Index")
plot(DATA$Dist_Water, main = "Distance to Water Sources")
```

```{r}
# Fitting classes to elevation data
elev_classes <- cut(DATA$Elevation,
                    3,
                    labels = c("low", "medium", "high"))
# Plotting classes with eagle occurrence overlay
plot(elev_classes, main = "Elevation Classes with Eagle Occurrences")
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'white', cex = 0.7)
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'black', cex = 0.5)

# Fitting classes to HFI data
hfi_classes <- cut(DATA$HFI,
                    3,
                    labels = c("low", "medium", "high"))
plot(hfi_classes, main = "HFI Classes with Eagle Occurrences")
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'white', cex = 0.7)
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'black', cex = 0.5)

# Fitting classes to Dist_Water data
water_classes <- cut(DATA$Dist_Water,
                    3,
                    labels = c("low", "medium", "high"))
plot(water_classes, main = "Distance to Water Classes with Eagle Occurrences")
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'white', cex = 0.7)
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'black', cex = 0.5)

# Fitting classes to Forest data
forest_classes <- cut(DATA$Forest,
                    3,
                    labels = c("low", "medium", "high"))
plot(forest_classes, main = "Forest Coverage Classes with Eagle Occurrences")
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'white', cex = 0.7)
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'black', cex = 0.5)
```

It is difficult to discern the relationship between eagle occurrences and distance to water as well as HFI. There is some suggestion that the eagles tend to stray away from elevation, mostly staying within the 'low' to 'medium' elevation areas. Additionally, for forested areas, we can see that the eagle occurrences typically are located in 'high' areas of forest coverage. 

## 2) Exploratory Data Analysis

We will do some first moment descriptive statistics - obtaining the intensity ($\lambda$) of eagle occurrences. 

```{r}
intensity <- intensity(eagles_ppp)
win_km <- rescale(Window(eagles_ppp), 1000, "km") # Rescale the window to units of km
npoints(eagles_ppp)/area(win_km) # Intensity in parks/km^2
```

The intensity of eagle occurrences per km$^2$ in BC is 0.005272802 eagles per km (under the assumption of homogeneity). Based on the previous visualizations, we suspect that there is spatial variation of the eagle occurrences ($\lambda$ is spatially varying), so we will estimate the intensity in each quadrat by counting the number of eagle occurrences in each quadrat divided by the quadrat's area. This will help us determine if the assumption of homogeneity is met. 

```{r}
# Split into a 10 by 10 quadrat and count points, then perform the quadrat test
Q <- quadratcount(eagles_ppp,
                  nx = 10,
                  ny = 10)
quadrat.test(Q)

# Visualizing both the quadrats and estimated intensity, with points
plot(Q, main = "Quadrat counts and eagle occurrence points")
points(eagles_ppp, pch = 20, col = 'red', cex = 0.5) 

intensity_map <- density(eagles_ppp)
plot(intensity_map, main = "Estimated intensity and eagle occurrence points")
points(eagles_ppp, pch = 20, col = 'white', cex = 0.7) 
points(eagles_ppp, pch = 20, col = 'black', cex = 0.5) 
```

We can see that the assumption of homogeneity does not hold for this dataset as the eagle occurrences tend to be clustered in some areas (like the southern coast area) whereas some areas have none at all. The quadrat counting suggests a spatially varying, inhomogeneous $\lambda(u)$. Note that we do expect some variation by chance alone due to point processes being stochastic. The p-value resulting from testing for significant deviations is significant, indicating there there *is* significant deviation from homogeneity.

Next, we'll use kernel estimation to also estimate spatially varying $\lambda(u)$.

```{r}
# Estimate the intensity with likelihood cross-validation for bandwidth selection and plotting
lambda_u_hat <- density(eagles_ppp, adjust=1, bw="likelihood")
plot(lambda_u_hat, main = "Kernel estimate of Eagle Occurence intensity")
plot(eagles_ppp, pch = 16, cex = 0.7, col = "white", add = TRUE)
plot(eagles_ppp, pch = 16, cex = 0.5, col = "black", add = TRUE)
```

The plot looks similar to that of the estimated intensity without the kernel estimation.

We can do a hot spot analysis to identify areas of elevated intensity to see where the eagle occurrences are concentrated.

```{r}
# Performing the hotspot analysis
R <- bw.ppl(eagles_ppp) # Estimate bandwidth for radius for scan test
LR <- scanLRTS(eagles_ppp, r = R) # Calculate the likelihood ratio test statistic
plot(LR, clipwin = eagles_ppp$window, main = "Likelihood Ratio Test Statistic for Eagle Occurrences") # Plot the test statistic

# Compute and plotting local p-values
pvals <- eval.im(pchisq(LR, df = 1, lower.tail = FALSE))
plot(pvals, main = "Local p-values for BC Parks", clipwin = eagles_ppp$window)
```

Interestingly, we do not see any particular 'hot spots' for eagle occurrences. When showing the local p-values, we do some scattered points with very low p-values that visibly seem to be around the lower half and within the interior of the BC. 

To investigate whether the intensity depends on the covariates, we will estimate $\rho$.

```{r, warning=F}
# Extracting the covariate information
elevation <- DATA$Elevation
water <- DATA$Dist_Water
forest <- DATA$Forest
hfi <- DATA$HFI

# Estimate Rho and plotting
rho_elev <- rhohat(eagles_ppp, elevation)
plot(rho_elev, xlab="Elevation (m)", ylab="Intensity (rho)", main="Rho vs. Elevation", xlim=c(0, max(rho_elev$elevation)))

rho_water <- rhohat(eagles_ppp, water)
plot(rho_water, xlab="Distance to Water (m)", ylab="Intensity (rho)", main="Rho vs. Distance to Water", xlim=c(0, max(rho_water$water)))

rho_forest <- rhohat(eagles_ppp, forest)
plot(rho_forest, xlab="Forest Coverage (%)", ylab="Intensity (rho)", main="Rho vs. Forest Coverage", xlim=c(0, max(rho_forest$forest)))

rho_hfi <- rhohat(eagles_ppp, hfi)
plot(rho_hfi, xlab="HFI", ylab="Intensity (rho)", main="Rho vs. HFI", xlim=c(0, max(rho_hfi$hfi)))
```

From the plots, we can see that eagles don't seem to stay around high elevated areas and they typically prefer lower forest coverage (ie. 20%). Interestingly, we can see that there is some elevated intensity of eagle occurrences in areas of higher HFI. As for distance to water, there isn't any notable evidence on whether eagles prefer to be closer to water or not, it seems as if they are rather spread out in areas close and not so close to water. 

We will now proceed by doing second moment statistics. We want to investigate the relationship between points.

We will use the Ripley's K-function to quantify the spatial dependence within our point pattern dataset to describe how the average number of additional events within a certain distance of a typical event compares to what would be expected under complete spatial randomness (CSR). 

```{r}
# Using Ripley's K-function and testing for a significant correlation
k_parks <- Kest(eagles_ppp, correction="border")
k_parks # Print summary

# For alpha = 0.05
E_eagles <- envelope(eagles_ppp, Kest, correction="border", nsim=19, verbose=FALSE, savefuns=TRUE)
plot(E_eagles, main="", lwd=2)

# Estimate a strictly positive density
lambda_eagles_pos <- density(eagles_ppp,
                          sigma=bw.ppl,
                          positive=TRUE)

# Simulation envelope (with points drawn from the estimated intensity)
E_eagles_inhom <- envelope(eagles_ppp,
                        Kinhom,
                        simulate = expression(rpoispp(lambda_eagles_pos)),
                        correction="border",
                        rank = 1,
                        nsim = 19,
                        fix.n = TRUE)

# visualise the results
par(mfrow = c(1,2))
plot(E_eagles_inhom,
     main = "",
     lwd = 2)

# Zoom in on range where significant deviations appear
plot(E_eagles_inhom,
     xlim = c(0, 150000),
     main = "",
     lwd = 2)
```

The black line (empirical $K$-function) devaites significantly from the grey area and the theoretical red line. This suggests that there is a departure from Complete Spatial Randomness, indicating that there is spatial correlation at certain scales (at some eagle occurrences locations). The estimators still assume homogeneity, which we relaxed via the `Kinhom()` function.

When corrected for homogeneity, we don't see evidence of clustering as the black line still seems to stay within the shaded area. Between 0 m and 75000 m, there seems to be some slight potential for clustering as the black line seems to lay right along the border of the shaded area, but it may not be enough to be significant. In other words, we can deduce that there is no evidence of clustering of eagle occurrences based on these plots. 

We will move on with a pair correlation function which gives localized information about the spatial structure at exactly a certain distance.

```{r}
# Estimate a strictly positive density for inhomogeneous correction
lambda_eagles_pos <- density(eagles_ppp, sigma=bw.ppl(eagles_ppp), positive=TRUE) 

# Simulation-based envelope test for the inhomogeneous and homogenous pair correlation function
pcf_eagles_inhom_env <- envelope(eagles_ppp,
                          fun=pcfinhom,
                          simulate = expression(rpoispp(lambda_eagles_pos)),
                          rank = 1,
                          nsim = 19)

pcf_eagles_hom_env <- envelope(eagles_ppp,
                          fun=pcf,
                          rank = 1,
                          nsim = 19)

# Visualizing the results
plot(pcf_eagles_hom_env, main="Homogeneous PCF with Simulation Envelope for Eagle Occurrences")
plot(pcf_eagles_inhom_env, main="Inhomogeneous PCF with Simulation Envelope for Eagle Occurrences")
```



## 3) Fitting Poisson point process model on the eagle point patterns

Since from the analysis of first moment descriptive statistics on bald eagle occurrences, we found that the intensity $\lambda(u)$ is inhomogeneous, which indicates $\lambda(u)$ is not constant in space but rather a function of some covariate(s).

Now we will model $\rho$ with the covariates, we will first make sure there is no colinearity between the covariates that will cause poor estimates of the regression coefficients. 

```{r}
cor.im(DATA[2:5], use = "pairwise.complete.obs")
```
The correlation coeffient between each pair of covariates are relatively weak, therefore, there is not too much concern on colinearity.

Before fitting a poisson process model on the data, we will begin by examining the relationship between the covariates and $\lambda(u)$ (number of bald eagles per km$^2$).

From the estimate of $\rho$ in the analysis of first moment descriptive statistics, we found that elevation, HFI and forest coverage exhibit a quadratic relationship with the number of bald eagles. And there isn't relationship between distance to water and the number of bald eagles.

So our initial guess of the model would be:
$$\lambda_{Eagle}(u)=e^{\beta_0+\beta_1\ elevation(u)+\beta_2\ elevation(u)^2+\beta_3\ forest\ cover(u)+\beta_4\ forest\ cover(u)^2+\beta_5\ HFI(u)+\beta_6\ HFI(u)^2}$$
```{r}
fit <- ppm(eagles_ppp ~ Elevation + I(Elevation^2) + Forest + I(Forest^2) + HFI + I(HFI^2), data = DATA)
fitnull <- ppm(eagles_ppp, data = DATA)

```
AIC/ANOVA

Next, we conduct an anova as well as find AIC values for the null fit and the initial guess fit.

```{r}
anova(fitnull, fit, test = 'LR')
AIC(fitnull) - AIC(fit)
```

Based on these values, we see that 1) the Anova shows that the increased complexity is warranted. Additionally 2) the AIC is positive which indicates that the fit null had a higher AIC than the initial guess fit. Both of these things indicate that we have chosen a fit that is at least better than a null fit.

To determine next whether this is optimally capturing the patterns of the covariates appropriately, we should visualize the partial residuals of each component.

```{r}
par_res_elev <- parres(fit, "Elevation")
par_res_forest <- parres(fit, "Forest")
par_res_hfi <- parres(fit, "HFI")

par(mfrow = c(2,2))
plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Elevation (m)")
plot(par_res_forest,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Forest Coverage (%)")
plot(par_res_hfi,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "HFI Levels")


```

Based on these initial partial residual plots, we can see that our initial guess was fairly decent for HFI levels, but we can clearly see that there is some complex patterns that are not quite being captured by the model for Elevation and Fores Coverage.

Since these appear to be a bit more complex relationships, an appropriate measure could be to try to use splines with some degrees of freedom. Again, we should check the resulting ANOVA and compare the new model with splines to the initial guess, and also find the AIC.

$$
\lambda_{\text{Eagle}}(u) = \exp\left(\beta_0 + f_1(\text{elevation}(u)) + f_2(\text{forest cover}(u)) + \beta_5 HFI(u) + \beta_6 HFI(u)^2\right)
$$


```{r}

library(splines)

# Testing  Model Complexity
fit2 <- ppm(eagles_ppp ~ Elevation * (Elevation^2) *bs(Elevation, df=20) + Forest * bs(Forest, df= 10) + HFI + I(HFI^2), data = DATA, use.gam = TRUE)

anova(fit, fit2, test = "LR")
AIC(fit) - AIC(fit2)

par_res_elev <- parres(fit2, "Elevation")
par_res_forest <- parres(fit2, "Forest")
par_res_hfi <- parres(fit2, "HFI")

par(mfrow = c(2,2))
plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Elevation (m)")
plot(par_res_forest,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Forest Coverage (%)")
plot(par_res_hfi,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "HFI Levels")

```
There is a lot to unpack here, but the key takeaways are:

1) We see from the ANOVA that the added complexity incorporating splines was significant and thus warranted.

2) we see that the difference in AIC between the initial guess and the spline guess was positive, indicating that the added complexity using splines was a more optimal fit to the data.

3) Based on the partial residuals plot, we can see that this tends to better capture the trends seen in the data (at least visually). We see more of an overlap of the dashed line and the curve. This indicates that the patterns are being captured well. Now, it is important to note that although this is better than the initial guess, there may be some room for improvement. However, this is already looking fairly reasonable. The choice of degrees of freedom (df) were set fairly high to allow for increased flexibility in the fitting process. I did tinker with this a bit to try and find an optimal fit.

With that being said, we can now go on to plot the actual model fit.

Plotting the Model Fit compared to the distribution from our ppp object

```{r}
plot(fit2,
     se = FALSE,
     superimpose = FALSE,
     log = TRUE,
     n = 1000)
plot(eagles_ppp,
     cols = 4,
     pch = 16,
     main = "Eagles occurrence data in BC",
     cex = 0.5)
```
Just based off initial visual inspection, it appears to be doing a decent job at modelling the ppp object which is expected since the partial residuals was fit, generally, quite well.

Quadrat Test

```{r}
quadrat.test(fit, nx = 2, ny = 4)

```
The small p value indicates that the data  significant deviates from the model...This test indicates that there is definitely some room for improving the model fit. However, it is important to note that this is subject to choices like quadrat size.


But let's take a closer look at the residuals. 

Looking at the Residuals

```{r}
#Calculate the residuals
res <- residuals(fit2, drop = TRUE)

plot(res,
     cols = "transparent",
     main = "Residuals of ppm")
```
Here, the overall residuals indicate that, particularly near the southern coast, the model significnatly struggles in predicting the intensity of eagles. In other areas, it appears to be performing fairly well, but there are some spot instances of poorer performance. Overall, however, this is a decent performing model.

The fit using the splines and quadratic terms indicates that the distribution of eagles is correlated to elevation, forests, and HFI in a non-linear fashion. We also found that distance to water had no relationship to the intensity of eagles. To illustrate this, consider the below:



```{r}

# Testing  Model Complexity
fit3 <-  ppm(eagles_ppp ~ Elevation * (Elevation^2) *bs(Elevation, df=20) + Forest * bs(Forest, df= 10) + HFI + I(HFI^2) + Dist_Water, data = DATA, use.gam = TRUE)

par_res_elev <- parres(fit3, "Elevation")
par_res_forest <- parres(fit3, "Forest")
par_res_hfi <- parres(fit3, "HFI")
par_res_water <- parres(fit3, "Dist_Water")


par(mfrow = c(2,2))
plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Elevation (m)")
plot(par_res_forest,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Forest Coverage (%)")
plot(par_res_hfi,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "HFI Levels")
plot(par_res_water,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Distance to Water")
```
The key illustration here is that water is completely flat, and thus is not correlated whatsoever to the intensity of Eagles. Thus, we should simply choose to omit this covariate from the model, and just use fit2.
