---
title: "Bald eagle in British Columbia - Spatial Exploration"
author: "Shayla Tran, Matthew Angoh & Kyle Deng"
date: "2024-04-17"
output: html_document
---

There is a total of 104,941 occurrences recorded of Bald eagle or *Haliaeetus leucocephalus (Linnaeus, 1766)* within Canada between 1848 and 2024. This data is directly downloaded from the source GBIF.

## 1) Initial Data Inspection and Data Exploration

Here, we load in the `BC_Covariates.Rda` data. We will also read in the `eagles_bc.csv` dataset which contains 5000 sample occurrences of Bold eagle.

```{r, include=FALSE, message=FALSE}
# Load necessary libraries
library(tidyverse)
library(spatstat)
library(sf)
library(polyCub)
```

```{r, warning=FALSE, message=FALSE}
load("data/BC_Covariates.Rda") # BC covariate data with window
occurrence_eagles <- read_csv("data/eagles_bc.csv") # Reading in the occurrences data
```

Creating the `ppp` object and plotting the Bald eagle occurrences into the BC window.

```{r}
window <- as.owin(DATA$Window) # Create an owin object with the window in the covariates dataset so it can be used to create the ppp object later

# Create the ppp object
eagles_ppp <- ppp(x = occurrence_eagles$X, y = occurrence_eagles$Y, window = window) # Obtain a ppp object with the coordinate and window data 

# Plot the ppp object
plot(eagles_ppp,
     cols = 4,
     pch = 16,
     main = "Eagles occurrence data in BC",
     cex = 0.5)
```

From the initial plot, we can see that the eagles tend to be near the southern regions and there are some densities along the coast and within the interior region.

The BC covariates provided are:

* HFI (Human Foot Index): influence of humans on the land
* Elevation: land elevation
* Forest: forested areas
* Dist_water: distance to water sources or bodies of water

To explore these covariates, we will create plots that show the eagle occurrences along with the covariate data within the window. Each covariate image will be split into 3 elevation classes (for now).

```{r, hide=TRUE}
# Plotting the covariates (for comparisons later)
plot(DATA$Elevation, main = "Elevation")
plot(DATA$Forest, main = "Forest Coverage")
plot(DATA$HFI, main = "Human Foot Index")
plot(DATA$Dist_Water, main = "Distance to Water Sources")
```

```{r}
# Fitting classes to elevation data
elev_classes <- cut(DATA$Elevation,
                    3,
                    labels = c("low", "medium", "high"))
# Plotting classes with eagle occurrence overlay
plot(elev_classes, main = "Elevation Classes with Eagle Occurrences")
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'white', cex = 0.7)
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'black', cex = 0.5)

# Fitting classes to HFI data
hfi_classes <- cut(DATA$HFI,
                    3,
                    labels = c("low", "medium", "high"))
plot(hfi_classes, main = "HFI Classes with Eagle Occurrences")
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'white', cex = 0.7)
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'black', cex = 0.5)

# Fitting classes to Dist_Water data
water_classes <- cut(DATA$Dist_Water,
                    3,
                    labels = c("low", "medium", "high"))
plot(water_classes, main = "Distance to Water Classes with Eagle Occurrences")
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'white', cex = 0.7)
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'black', cex = 0.5)

# Fitting classes to Forest data
forest_classes <- cut(DATA$Forest,
                    3,
                    labels = c("low", "medium", "high"))
plot(forest_classes, main = "Forest Coverage Classes with Eagle Occurrences")
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'white', cex = 0.7)
points(eagles_ppp$x, eagles_ppp$y, pch = 20, col = 'black', cex = 0.5)
```

It is difficult to discern the relationship between eagle occurrences and distance to water as well as HFI. There is some suggestion that the eagles tend to stray away from elevation, mostly staying within the 'low' to 'medium' elevation areas. Additionally, for forested areas, we can see that the eagle occurrences typically are located in 'high' areas of forest coverage. 

2) Exploratory Data Analysis

We will do some first moment descriptive statistics - obtaining the intensity ($\lambda$) of eagle occurrences. 

```{r}
intensity <- intensity(eagles_ppp)
win_km <- rescale(Window(eagles_ppp), 1000, "km") # Rescale the window to units of km
npoints(eagles_ppp)/area(win_km) # Intensity in parks/km^2
```

The intensity of eagle occurrences per km$^2$ in BC is 0.005272802 eagles per km (under the assumption of homogeneity). Based on the previous visualizations, we suspect that there is spatial variation of the eagle occurrences ($\lambda$ is spatially varying), so we will estimate the intensity in each quadrat by counting the number of eagle occurrences in each quadrat divided by the quadrat's area. This will help us determine if the assumption of homogeneity is met. 

```{r}
# Split into a 10 by 10 quadrat and count points, then perform the quadrat test
Q <- quadratcount(eagles_ppp,
                  nx = 10,
                  ny = 10)
quadrat.test(Q)

# Visualizing both the quadrats and estimated intensity, with points
plot(Q, main = "Quadrat counts and eagle occurrence points")
points(eagles_ppp, pch = 20, col = 'red', cex = 0.5) 

intensity_map <- density(eagles_ppp)
plot(intensity_map, main = "Estimated intensity and eagle occurrence points")
points(eagles_ppp, pch = 20, col = 'white', cex = 0.7) 
points(eagles_ppp, pch = 20, col = 'black', cex = 0.5) 
```

We can see that the assumption of homogeneity does not hold for this dataset as the eagle occurrences tend to be clustered in some areas (like the southern coast area) whereas some areas have none at all. The quadrat counting suggests a spatially varying, inhomogeneous $\lambda(u)$. Note that we do expect some variation by chance alone due to point processes being stochastic. The p-value resulting from testing for significant deviations is significant, indicating there there *is* significant deviation from homogeneity.

Next, we'll use kernel estimation to also estimate spatially varying $\lambda(u)$.

```{r}
# Estimate the intensity with likelihood cross-validation for bandwidth selection and plotting
lambda_u_hat <- density(eagles_ppp, adjust=1, bw="likelihood")
plot(lambda_u_hat, main = "Kernel estimate of Eagle Occurence intensity")
plot(eagles_ppp, pch = 16, cex = 0.7, col = "white", add = TRUE)
plot(eagles_ppp, pch = 16, cex = 0.5, col = "black", add = TRUE)
```

The plot looks similar to that of the estimated intensity without the kernel estimation.

We can do a hot spot analysis to identify areas of elevated intensity to see where the eagle occurrences are concentrated.

```{r}
# Performing the hotspot analysis
R <- bw.ppl(eagles_ppp) # Estimate bandwidth for radius for scan test
LR <- scanLRTS(eagles_ppp, r = R) # Calculate the likelihood ratio test statistic
plot(LR, clipwin = eagles_ppp$window, main = "Likelihood Ratio Test Statistic for Eagle Occurrences") # Plot the test statistic

# Compute and plotting local p-values
pvals <- eval.im(pchisq(LR, df = 1, lower.tail = FALSE))
plot(pvals, main = "Local p-values for BC Parks", clipwin = eagles_ppp$window)
```

Interestingly, we do not see any particular 'hot spots' for eagle occurrences. When showing the local p-values, we do some scattered points with very low p-values that visibly seem to be around the lower half and within the interior of the BC. 

To investigate whether the intensity depends on the covariates, we will estimate $\rho$.

```{r}
# Extracting the covariate information
elevation <- DATA$Elevation
water <- DATA$Dist_Water
forest <- DATA$Forest
hfi <- DATA$HFI

# Estimate Rho and plotting
rho_elev <- rhohat(eagles_ppp, elevation)
plot(rho_elev, xlab="Elevation (m)", ylab="Intensity (rho)", main="Rho vs. Elevation", xlim=c(0, max(rho_elev$elevation)))

rho_water <- rhohat(eagles_ppp, water)
plot(rho_water, xlab="Distance to Water (m)", ylab="Intensity (rho)", main="Rho vs. Distance to Water", xlim=c(0, max(rho_water$water)))

rho_forest <- rhohat(eagles_ppp, forest)
plot(rho_forest, xlab="Forest Coverage (%)", ylab="Intensity (rho)", main="Rho vs. Forest Coverage", xlim=c(0, max(rho_forest$forest)))

rho_hfi <- rhohat(eagles_ppp, hfi)
plot(rho_hfi, xlab="HFI", ylab="Intensity (rho)", main="Rho vs. HFI", xlim=c(0, max(rho_hfi$hfi)))
```

From the plots, we can see that eagles don't seem to stay around high elevated areas and they typically prefer lower forest coverage (ie. 20%). Interestingly, we can see that there is some elevated intensity of eagle occurrences in areas of higher HFI. As for distance to water, there isn't any notable evidence on whether eagles prefer to be closer to water or not, it seems as if they are rather spread out in areas close and not so close to water. 

We will now proceed by doing second moment statistics. We want to investigate the relationship between points.

We will use the Ripley's K-function to quantify the spatial dependence within our point pattern dataset to describe how the average number of additional events within a certain distance of a typical event compares to what would be expected under complete spatial randomness (CSR). 

```{r}
# Using Ripley's K-function and testing for a significant correlation
k_parks <- Kest(eagles_ppp, correction="border")
k_parks # Print summary

# For alpha = 0.05
E_eagles <- envelope(eagles_ppp, Kest, correction="border", nsim=19, verbose=FALSE, savefuns=TRUE)
plot(E_eagles, main="", lwd=2)

# Estimate a strictly positive density
lambda_eagles_pos <- density(eagles_ppp,
                          sigma=bw.ppl,
                          positive=TRUE)

# Simulation envelope (with points drawn from the estimated intensity)
E_eagles_inhom <- envelope(eagles_ppp,
                        Kinhom,
                        simulate = expression(rpoispp(lambda_eagles_pos)),
                        correction="border",
                        rank = 1,
                        nsim = 19,
                        fix.n = TRUE)

# visualise the results
par(mfrow = c(1,2))
plot(E_eagles_inhom,
     main = "",
     lwd = 2)

# Zoom in on range where significant deviations appear
plot(E_eagles_inhom,
     xlim = c(0, 150000),
     main = "",
     lwd = 2)
```

The black line (empirical $K$-function) devaites significantly from the grey area and the theoretical red line. This suggests that there is a departure from Complete Spatial Randomness, indicating that there is spatial correlation at certain scales (at some eagle occurrences locations). The estimators still assume homogeneity, which we relaxed via the `Kinhom()` function.

When corrected for homogeneity, we don't see evidence of clustering as the black line still seems to stay within the shaded area. Between 0 m and 75000 m, there seems to be some slight potential for clustering as the black line seems to lay right along the border of the shaded area, but it may not be enough to be significant. In other words, we can deduce that there is no evidence of clustering of eagle occurrences based on these plots. 

We will move on with a pair correlation function which gives localized information about the spatial structure at exactly a certain distance.

```{r}
# Estimate a strictly positive density for inhomogeneous correction
lambda_eagles_pos <- density(eagles_ppp, sigma=bw.ppl(eagles_ppp), positive=TRUE) 

# Simulation-based envelope test for the inhomogeneous and homogenous pair correlation function
pcf_eagles_inhom_env <- envelope(eagles_ppp,
                          fun=pcfinhom,
                          simulate = expression(rpoispp(lambda_eagles_pos)),
                          rank = 1,
                          nsim = 19)

pcf_eagles_hom_env <- envelope(eagles_ppp,
                          fun=pcf,
                          rank = 1,
                          nsim = 19)

# Visualizing the results
plot(pcf_eagles_hom_env, main="Homogeneous PCF with Simulation Envelope for Eagle Occurrences")
plot(pcf_eagles_inhom_env, main="Inhomogeneous PCF with Simulation Envelope for Eagle Occurrences")
```

